From Reversible Programs to \\ Univalent Universes and Back

### 2017-05-04 upi.tex

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
and the extension with recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}. In this paper we will
identify a ``featherweight'' representative to use for the formal
development. But first, we illustrate the flavor of the family of languages with
some examples written in a Haskell syntax.

First consider the following program:

{\footnotesize
\begin{verbatim}
addSub1 :: (Int, Int) :<=> (Int, Int)
addSub1 = CommuteTimes
      :.: (UnfoldN :*: Id)
      :.: Distribute
      :.: (Id :+: CommuteTimes)
      :.: Factor
      :.: (FoldN :*: Id)
\end{verbatim}}
%
which takes a pair of natural numbers and processes them as follows depending on
whether the second number is 0 or not. If the second number is 0, the evaluation
proceeds as follows:
$(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
\mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
$(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
\mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
(\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
applications of this program would increment the first number and decrement the
second number maintaining the sum at all times. When the second number reaches
0, the numbers are swapped.

As another example, consider the following expressions culminating with the
implementation of the Toffoli gate, whichi is universal for combinational
circuits. We have two implementations of simple boolean negation: a direct one
and a more convoluted one. These two implementations are ``equivalent'' in the
sense to be explained below. The expression \verb|cond| is the main reversible
control flow mechanism: it allows for the conditional execution of one of two
branches but retains the choice of which branch. This expression can be used
twice to implement the Toffoli gate, which performs a conditional negation on
one bit if two other bits are both true:

@@

{\footnotesize
\begin{verbatim}
negOpt :: inot2 :<===> inot
negOpt = (Id :>: AssocLeftEquiv)
    :..: (Id :>: (SlideCommuteTimes :>: Id))
    :..: (Id :>: AssocRightEquiv)
    :..: (Id :>: (Id :>: AssocLeftEquiv))
    :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
    :..: (Id :>: (Id :>: IdElimL))
    :..: AssocLeftEquiv
    :..: (SlideUnitI :>: Id)
    :..: AssocRightEquiv
    :..: (Id :>: InvInvEquivId)
    :..: IdElimR
\end{verbatim}}

This program, which can be thought of as an optimization or a proof, relates the
two implementations of boolean negation via a series of elementary
optimizations of which the following are the most interesting:

@@

\begin{verbatim}

conjecture it is related to HoTT

study that language and formalize the connections between its combinators and
between HoTT

stud the smallest core that can explain the idea: restrict to just B; essence
not losing anything really
\end{verbatim}

@@

%%%%%
\subsection{The Language $\Pi$ of Type Isomorphisms}

The 2-combinators $\alpha$ also have inverses. The above forms a 2-groupoid (see
accompanying code once it is fixed).

>>>>

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
and the extension with recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}. In this paper we will
identify a ``featherweight'' representative to use for the formal
development. But first, we illustrate the flavor of the family of languages with
some examples written in a Haskell syntax. First consider the following program:

{\footnotesize
\begin{verbatim}
addSub1 :: (Int, Int) :<=> (Int, Int)
addSub1 = CommuteTimes
      :.: (UnfoldN :*: Id)
      :.: Distribute
      :.: (Id :+: CommuteTimes)
      :.: Factor
      :.: (FoldN :*: Id)
\end{verbatim}}
%
\noindent which takes a pair of natural numbers and processes them as follows depending on
whether the second number is 0 or not. If the second number is 0, the evaluation
proceeds as follows:
$(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
\mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
$(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
\mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
(\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
applications of this program would increment the first number and decrement the
second number maintaining the sum at all times. When the second number reaches
0, the numbers are swapped.

As another example, consider the following expressions culminating with the
implementation of the Toffoli gate, which is universal for combinational
circuits. We have two implementations of simple boolean negation: a direct one
and a more convoluted one. These two implementations are ``equivalent'' in the
sense to be explained below. The expression \verb|cond| is the main reversible
control flow mechanism: it allows for the conditional execution of one of two
branches but retains the choice of which branch was executed. This mechanism can
be used twice to implement the Toffoli gate, which performs a conditional
negation on one bit if two other bits are both true:

@@

{\footnotesize
\begin{verbatim}
negOpt :: inot2 :<===> inot
negOpt = (Id :>: AssocLeftEquiv)
    :..: (Id :>: (SlideCommuteTimes :>: Id))
    :..: (Id :>: AssocRightEquiv)
    :..: (Id :>: (Id :>: AssocLeftEquiv))
    :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
    :..: (Id :>: (Id :>: IdElimL))
    :..: AssocLeftEquiv
    :..: (SlideUnitI :>: Id)
    :..: AssocRightEquiv
    :..: (Id :>: InvInvEquivId)
    :..: IdElimR
\end{verbatim}}
%
\noindent This program, which can be thought of as an optimization or a proof, relates the
two implementations of boolean negation via a series of elementary
optimizations of which the following are the most interesting:

@@

In previous work, on $\Pi$ we noted a possible connection with HoTT:
\begin{quote}
  It is, therefore, at least plausible that a variant of HoTT based exclusively
  on reversible functions that directly correspond to equivalences would have
  better computational properties. Our current result is a step, albeit
  preliminary, in that direction as it only applies to finite
  types~\cite{Carette2016}.
\end{quote}
Formalizing, in a precise sense, the connection between reversible programs
based on combinators and paths in HoTT, as intuitive as it may seem, is however
difficult. Paths in HoTT come equipped with principles like the
``contractibility of singletons'', ``transport'', and ``path induction.'' None
of these principles seems to have any direct counterpart in the world of
reversible programming.

We will however show, in a very clear sense, how a $\Pi$-based language is
``equivalent'' to a univalent HoTT universe that possesses all the required
properties of paths. The essence of the construction can be demonstrated with a
tiny ``featherweight'' variant of $\Pi$ that includes just one type: booleans,
and a tiny ``featherweight'' univalent universe to be formalized in the next
section. Here we give a formalization of the small $\Pi$-based language which we
will use in the formalization in the rest of the paper.

The 2-combinators $\alpha$ also have inverses. The above forms a 2-groupoid.

@@
